### 2025-04-09 chapters/3-jets.tex
T-1 day to initial submission, nearly complete draft up to citations and small comments.

\epigraph{Never trust an parton: they make up everything. Even themselves!}{Joke of the author}

@@

In this chapter, we get a better handle on things by giving up on thinking about individual particles.
%
We instead make progress by thinking about many particles at once!
%
These groups of particles are called \emph{jets},\footnote{\sam{roughly}} and we will address them theoretically with a common strategy in quantum mechanics:
%
summing up all of the different possibilities.
%
We consider all the ways a particle could split, how each of its children could split, and so on, and add together the probabilities of each to obtain a consistent estimate of the behavior of jets.
%
This chapter initiates the study of jets and jet substructure, which we define shortly, and introduces some of the key conceptual and computational tools used in the work presented in this thesis.

@@

We will be using the building blocks of particles and splitting, where we have already seen the use of \glslink{regularization}{\underline{regularization}} -- the removal of infinities to produce a physically sensible result.
%
We will use \glslink{resummation}{\underline{resummation}} -- the use of series techniques and calculus to turn infinite series into analytic or numerical expressions -- to add up the results of an infinite number of parton-level splittings and uncover the secrets of jet substructure.
%
We will see \underline{renormalization} -- roughly, ``zooming in'' -- emerge as a useful lens as particles seem to split more and more as we zoom further and further in on jets.

% -----------------------------------
% Picturebook figure
% -----------------------------------
\reusefigure[ht]{picturebook_jets}

@@

\epigraph{The typical high-energy event should consist in the production of a pair of virtual partons, each of which develops into a jet of physical hadrons.}{N. Cabibbo, G. Parisi, and M. Testa, \textit{Hadron Production in \(e^+ e^-\) Collisions} \cite{}, 1970}
% https://link.springer.com/article/10.1007/BF02755392

\epigraph{Continuing in this way, it should be possible to test the detailed predictions of QCD for production of any numbers of quarks and gluons, but always reinterpreting
these particles in terms of jets.}{G. Sterman and S. Weinberg, \textit{Jets from Quantum Chromodynamics} \cite{}, 1977}

We have seen that particle splittings are universal in the collinear limit, and that \sam{particles split more and more as we zoom in more and more on the narrow core surrounding a parton}.
%
The particles emerging from these splittings can be grouped together into universal objects that we call jets.
%
As discussed in \Sec{confinement} of the previous chapter, we strongly believe that quarks and gluons are \textit{confined}, and cannot be directly observed.
%
Therefore, jets are a key feature of high-energy particle collisions that we must use to infer the presence of quarks and gluons.

%%
%Each splitting is governed by a Dokshitzer-Gribov-Lipatov-Altarelli-Parisi (DGLAP) splitting function \cite{Gribov:1972ri,Dokshitzer:1977sg,Altarelli:1977zs}, \(P_{\{j\} \leftarrow i}\), which desribes the pseudo-probability that a parton of type \(i\) (i.e. \(i \in \{\text{quark, anti-quark, gluon}\}\) in QCD) splits into a set of partonic children with types \(\{j\}\).
%%
%The splitting functions govern the behavior of amplitudes in the collinear limit -- when a subset of outgoing partons have nearly collinear momenta -- and are therefore important tools for resummation in the physics of QCD jets, which tend to be fairly narrow.

\begin{definitionbox}{Tree of Partonic Splittings (The Partonic Cascade)}{tree}
    In \gls{pqcd}, a jet is formed when an initiating parton emerging out of a hard interaction recursively splits, or \vocab{showers}, into partons of lower energy.
    %
    The resulting \glslink{tree}{\vocab{tree of partonic splittings}} is a theoretical construct of \gls{pqcd} which contains the full history and information of all of the splittings of a single parent parton, including all information involving the kinematics and partonic flavor of each intermediate parton.
    %
    The tree of partonic splittings is also called the \vocab{partonic cascade}.
    %
    See \Fig{}.
\end{definitionbox}

\begin{figure}[]
    \centering
    \includegraphics[width=0.8\textwidth]{example-image-a}
    \caption{
        \sam{Cartoon depicting partonic cascade}
    }
    \label{fig:partonic-cascade}
\end{figure}

In order to provide useful information about our microscopic universe, jets must be defined in a way which is experimentally robust, theoretically well-defined, and produces results which can be compared to calculations in perturbation theory.
%
The study of the \textit{substructure} of jets, through jet shapes, angular distributions, and energy profiles, can then be used to provide additional insights into the properties of the partons involved in high-energy collider processes.
%
Furthermore, understanding jet formation and fragmentation is essential for measurements of key processes, such as Higgs boson production, top quark pair production, and searches for new physics beyond the Standard Model.

% These partons are produced in hard scatterings and subsequently fragment into multiple hadrons, which are detected as a group of particles in the detector. The process of jet formation can be described by parton showers, where the primary parton radiates secondary partons, leading to further fragmentation. The jet itself is characterized by its energy, momentum, and the pattern of particles in the detector that form the jet.

%Jets are typically identified and reconstructed from the final-state particles detected in the collider experiments using various algorithms, such as the anti-$k_T$ algorithm.
%
%Jets are important observables for studying the dynamics of strong interactions, and their properties are used to measure the characteristics of the underlying parton, including its energy, flavor, and angular distribution.

@@

In this thesis, we use the \gls{akt} \cite{}, \gls{ca} \cite{}, and \gls{kt} \cite{} jet algorithms, which are all \vocab{iterative \gls{reclustering} algorithms}, or \vocab{sequential recombination algorithms}:
%
they act on sets of outgoing momenta by recursively combining pairs of particles together until the algorithm reaches a termination condition.
%
When the algorithm terminates, it yields a set of jets associated with the input particle collision event.

The jet algorithms we use in this work (and nearly all jet algorithms) depend on a \vocab{jet radius} \Rjet, which determines the angular separation of particles within a jet.
%
Roughly, \(\Rjet\) is the radius of a circle in angular coordinates which determines a cone in which the jet constituents must lie;
%
if two particles are separated by an angle greater than \(2 \Rjet\), they will (usually) be associated with different jets.

\remark{}{
    The use of narrow jets, with \(\Rjet \ll 1\), justifies the use of the collinear limit, which will ease the calculation of observables which capture the substructure of jets.%
In particular, the physics of narrow jets can be accurately characterized by using the splitting probabilities we discussed in \Sec{collinear-limit}.
    %
    Furthermore, the use of \textit{central} jets, with rapidity \(y \sim 0\), will allow us to trade transverse momenta (which are more easily measured in hadronic collisions) for energy fractions (which appear in the theoretical description of jets through \glspl{splittingfn}).
}

@@

%We have stated that useful jet definitions must be theoretically well-defined and robust to experimental noise.
%%
%For now, we state without proof that anti-\(k_t\) jets produced with a WTA algorithm are well-defined and robust.
%%
%We will elaborate further in \Sec{}, where we discuss sources of experimental noise -- such as the effects of experimental detectors -- and theoretical concerns -- such as IRC safety.

Now that we have defined jets, we want to find a way to leverage them to determine the fundamental properties of our universe.
%
In this thesis, we take the approach of studying the \textit{substructure} of jets.
%
The study of jet substructure has contributed to the deep connection between experimental and theoretical particle physics, with notable successes including precise determinations of the strong coupling constant \(\alpha_s\) and the on-going search for the precise value of the top quark mass \(m_t\) \cite{} \sam{non-EEC, then EEC}, whose value dictates the stability of the \glslink{sm}{Standard Model} vacuum and therefore the fate of the universe as we know it.

@@

\remark{}{
    The algorithmic formulation of \glspl{jet} was created in order to study jets as proxies of the quarks and gluons of QCD;
    \sam{Cite Sterman-Weinberg, and others?}
    %
    in the algorithmic formulation, a jet is a coarse-grained degree of freedom roughly associated with a quark and its associated collinear radiation.
    %
    We note briefly, however, a complementary point of view that jets may also be thought of as a collinear enhancement of the \gls{energy-flow} (as in \Eq{energy-flow}) in a scattering event \cite{}.
    \sam{Cite original Russian papers here?}
    %
    The latter idea is more connected to rigorous CFT methods \cite{} and may also be more intuitively connected to experimental results -- the energy flow \(\mathcal{E}(\hat{n})\) is roughly the energy passing through a detector at the angle \(\hat{n}\).
}

@@

If you zoom away from the constituents of a system (or towards them), you're likely to be at a renormalization fixed point.

@@

One of the first uses of renormalization was in the study of the presence and size of magnetic domains within magnetic materials, such as naturally-occurring iron.
%
Renormalization tells us how water molecules form snowflakes, and snowflakes form snow banks.
%
It is the theory of how structure and order forms from simplicity, and how disorder and chaos form from structure.

@@

\remark{}{
    The use of subjets within jets already hints at a fractal-like, self-similar structure of the partonic cascade.
    %
    An extremely concrete description of jets as fractals can be obtained using the technology of the \gls{emd};
    %
    see \Sec{} and \Reff{}.
}

If \gls{renormalization} is the physics, then \vocab{\gls{resummation}} is the math.
%
\sam{I do resummation seemingly without renormalization in \Sec{ll-substructure-diagrams} -- telegraph more in this section and make this more clear in the more precise discussion later}
%
Theoretically, we have seen that jets can be imagined as the result of an arbitrary number of partonic splittings -- as few as zero and as many as infinity -- generated off of a parent quark, gluon, or other parton.
%
Resummation is generally the study of how to produce finite results from infinite series of this type -- especially when additional regularization is needed to produce a finite result.
%
Come to us in our hour of hope, it is exactly the mathematical framework we need.

\begin{definitionbox}{Resummation}{resummation}
    \glslink{resummation}{\vocab{Resummation}} is any method uused to turn (potentially divergent) series expansion into compact and often analytic mathematical expressions without any summation.
    %
    For our purposes, resummation will refer to the process of systematically reintroducing terms that have been previously expanded or neglected in perturbative expansions, in order to account for important contributions at higher orders or at specific scales.

%     \vspace{7pt}
%     \hrule
%     \vspace{7pt}

%     In quantum field theory and high-energy physics, \vocab{resummation} is used to improve perturbative expansions by resumming certain classes of diagrams or series of terms that would otherwise be neglected in the standard perturbative approach. Often, this procedure is applied to logarithmic or large-angle divergences that arise in certain physical processes, particularly in scattering and jet formation. Resummation techniques are particularly important in cases where higher-order terms in the perturbative expansion are large or grow with the energy scale, and their inclusion leads to more accurate predictions.

%     The need for resummation arises when conventional perturbation theory fails to converge or produces unreasonably large corrections. By resumming the series, the perturbative expansion becomes more controlled, and the predictions for physical observables, such as cross sections, become more reliable. Some commonly used resummation techniques include the exponentiation of leading logarithmic terms, the use of effective field theories to capture large momentum transfer effects, and the application of the Borel transform to reorganize divergent series.
\end{definitionbox}

Resummation is the technical segue we need to go from the conceptual developments of jets and renormalization to a quantitative study of jet substructure:
%
it is \textit{compulsory} to remove the infinite number of infinities -- each infinity emerging from a partonic splitting, as we saw in \Sec{} in our theoretical collision laboratory of \(e^+ e^-\to\,\)hadrons -- and it comes not a moment too soon.
%
\sam{mention logarithmic infinities?}
%
% Resummation is often used to deal with large logarithmic terms that arise in high-energy processes, such as in soft and collinear radiation. These terms can grow large at high energies or small momentum transfers, leading to large corrections in perturbative calculations. By resumming these logarithmic contributions, one can achieve a more controlled and accurate result. For example, in deep inelastic scattering, the leading-order cross section is often dominated by large logarithms of the form \( \ln(Q^2/m^2) \), where \( Q \) is the energy scale of the process, and \( m \) is the mass of the particle. Resummation techniques exponentiate these large terms to improve the accuracy of predictions.
%
% The two main types of divergences that motivate resummation in high-energy physics are collinear and soft divergences. Collinear divergences occur when emitted particles move along the same direction as the incoming particle, leading to enhanced radiation. Soft divergences, on the other hand, arise when particles emit soft (low-energy) radiation. Both types of divergences are often handled using resummation techniques, as they can contribute significantly to the final result in high-energy collisions, such as in jet production or hadronic interactions. In these cases, resummation provides a systematic way to deal with large-angle and low-energy radiation contributions.

Soon, we show that resummation of the infinity of partonic splittings, combined with the principle of \vocab{angular ordering} that we describe shortly, reduces the seemingly massive complexity of the tree of partonic emissions into compact formulae.
%
In fact, we characterize the fractal-like, psychadelic inner life of jets with simple exponential functions!
%
We say that the associated resummed quantities \vocab{exponentiate}.
%
A simple application of renormalization will then reveal what is called the \vocab{DGLAP Equation} -- a simple, physically intuitive result of scaling which reveals the deep heart of jet physics.

@@

\remark{}{
    The mathematical theory of resummation goes far beyond the paltry introduction of this chapter.
    %
    Borel resummation \cite{} is particularly important for understanding the full formal structure of quantum field theory (appearing, for example, in the theory of renormalons and the non-perturbative structure).
    %
    Abel resummation, named after the mathematician whose quote appears at the beginning of this section, is another common resummation method for divergent series in mathematics.
    %
    For a more mathematical and complete presentation, I can recommend the classic mathematical textbook \sam{\underline{add}} by Hardy \cite{} and the classic physics articles of \Reffs{}{};
    %
    unfortunately, a full list of the many wonderful resources for resummation beyond these, is far beyond my scope or ability \cite{}.
    %
    \sam{complete}
}
% Hardy, Renormalons, Alien Calculus
% https://public.websites.umich.edu/~jpboyd/boydactaapplicreview.pdf
% https://arxiv.org/pdf/hep-ph/0510142

\sam{Want discussion of LL, NLL, etc.}

\sam{Then discussion of MLL}

\begin{definitionbox}{N\(^k\)LL Accuracy}{nkll}
    \glslink{accuracy}{N\(^k\)LL}

    The main focus of this thesis will be LL results in jet substructure.
\end{definitionbox}

% ==============================================
\section{The Partonic Cascade: A Universal Model of Jet Formation}
% ==============================================
\label{sec:partonshower}

Finally, we are ready to describe the properties of the entire tree of partonic splittings.
%
As promised, we will address this partonic cascade by sewing together the results of consecutive partonic splittings, leading us to several deep and quite simple conclusions about the fractal nature of jets.
%
The simple description we provide in this chapter strictly holds only to LL accuracy, but nonetheless facilitates a quantitative study of jet substructure.
%
First, however, we will need one more fixed-order result that dramatically simplifies the all-orders physics of jet formation:
%
\underline{the principle of angular ordering}.

% ---------------------------------------
\subsection{Everything On-Shell: Strong Ordering of Successive Splittings}
% ---------------------------------------
\label{sec:angular-ordering}

\epigraph{The picture then is one where matrix elements are likely to provide the better description of the main character of events, i.e. the topology of well separated jets, while parton showers should be better at describing the internal structure of these jets.}{J. Andr\'e and T. Sj\"ostrand, \textit{A matching of matrix elements and parton showers} \cite{}}
    % https://arxiv.org/pdf/hep-ph/9708390

% https://www.hep.phy.cam.ac.uk/theory/webber/QCDlect3.pdf#page=31
% https://indico.cern.ch/event/492098/contributions/1169736/attachments/1356309/2050123/QCD2.pdf#page=21

% IRC Safety figs:
% https://indico.cern.ch/event/492098/contributions/1169736/attachments/1356309/2050123/QCD2.pdf#page=47

% Nice introduction in original paper!!!!
% https://www.sciencedirect.com/science/article/pii/0550321389905750
% Combine with QCD and Collider Physics (more opaque)
% Need to get eikonal factor... Schwartz?

In the previous chapter, we saw the splitting function emerge in the collinear limit as a universal tool for describing the splitting of an off-shell internal line of a Feynman diagram into a pair of on-shell final state particles.
%
Ideally, we would like to use these same ingredients -- the off-shell to on-shell splitting functions -- as building blocks to construct the entire parton cascade.
%
However, when we want to imagine a jet as successive splittings, do not have the luxury of considering only off-shell-to-on-shell splittings:
%
the successive splittings of the jet include splittings whose \textit{children are also off-shell}.

In this section, we overcome this apparent difficulty to produce a consistent and elegant characterization of the partonic cascade using only the simple (off-shell-to-on-shell) splitting functions of the previous chapter.
%
In particular, we show that \textit{angular ordering} -- a quantum interference phenomenon in perturbative QCD which dictates that final-state on-shell splittings are significantly more narrow in angle than the splittings of internal lines -- allows us to treat the more complicated splittings of internal lines as if their children are on-shell.
%
We then show that angular ordering itself is consistent with the simple description of the partonic cascade in terms of on-shell splittings.

\textit{An elegant and consistent description of the partonic cascade can be produced in terms of the simple splitting functions of the previous chapter and the prescription of angular ordering.}

\sam{Nearly on shell, drop the extra factors}

\begin{itemize}
    \item
Strong energy ordering of the first emission by the eikonal limit (soft theorems)

    \item
Strong angular ordering of the first emission by coherence

    \item
Therefore, the next emission is nearly on-shell;
%
any corrections are of the form of powers of \(\theta_0/\theta_1\), which are miniscule by angular ordering

    \item
Drop these terms. Recursion and strong ordering
\end{itemize}

\cite{osti_4367332} % Chudakov

\cite{Mueller:1981ex} % First, fixed-order and conjectured logarithmically

\cite{Dokshitzer:1982fh,Webber:1983if,Marchesini:1983bm} % Elaboration and implementation

\cite{Ciafaloni:1984zr} % Fadeev-Kulish

\cite{Ellis:1996mzs} % QCD and Collider Physics

\sam{Angular-ordering can be deduced, at least approximately, from the fixed-order structure of pQCD amplitudes.}
%
Before justifying this, however, let us
\begin{definitionbox}{Angular Ordering}{angular-ordering}
    \glslink{angular-ordering}{\vocab{Angular ordering}} is a principle of pQCD which, in the context of the partonic cascade, states that each virtual parton splits into two children, and subsequent splittings occur at exponentially smaller angles, as presented schematically in \Fig{shower:cartoon:fragmentation} and its caption.
    %
    \sam{emissions get narrower and narrower as one moves down the tree of partonic splittings -- i.e. as one gets further and further from the original parent parton}.
\end{definitionbox}

\begin{figure}[]
    \centering
    \includegraphics[width=0.8\textwidth]{example-image-a}
    \caption{
        \sam{Cartoon depicting angular ordering}
    }
    \label{fig:angular-ordering}
\end{figure}

\remark{}{
    The definition above is the functional one for our purpose.
    %
    However, there is a more intuitive explanation of angular ordering, as we will see.
    %
    Roughly, angular ordering states that, that \sam{in order to probe more and more deeply into the structure of the jet, one needs radiation of higher and higher energy moving transversely to the jet}.
    %
    In this sense, it seems closely related to the principles we have already discussed:
    %
    probing the structure of a geode requires hitting it hard;
    %
    probing the structure of a proton requires hitting it \textit{very} hard;
    %
    the branching history of the partonic cascade can only be resolved by sufficiently high energy partons.
    %
    This phenomenon is called \vocab{coherent parton branching}~\cite{Collins:2011zzd}.
}

\begin{itemize}
    \item
To deduce the principle of angular ordering, we will look at the amplitude \sam{obtained by... sewing together splitting functions? otherwise? Steal derivation}

>>>>

\epigraph{Never trust an parton: they make up everything. Even themselves!}{Sam}

@@

In this chapter, we get a better handle on things by giving up on thinking about individual particles.
%
We instead make progress by thinking about many particles at once!
%
These groups of particles are called \emph{jets}, and we will address them theoretically with a common strategy in quantum mechanics:
%
summing up all of the different possibilities.
%
We consider all the ways a particle could split, how each of its children could split, and so on, and add together the probabilities of each to obtain a consistent estimate of the behavior of jets.
%
This chapter initiates the study of jets and jet substructure, which we define shortly, and introduces some of the key conceptual and computational tools used in the work presented in this thesis.
%
If the exposition of this chapter is confusing or insufficient, I can highly recommend the pedagogical treatment of \Reff{Marzani:2019hun}.

@@

We will be using the building blocks of \Chap{particles};
%
particles remain the main actors of this section, and the substructure diagrams of \Sec{} will still be useful tools.
%
Most importantly, however, we will construct a model of jets using splitting functions, soft theorems and the angular-ordered principle of collinear color coherence.
%
% -- where we have already seen the use of \glslink{regularization}{\underline{regularization}} -- the removal of infinities to produce a physically sensible result.
%
We will use \glslink{resummation}{\underline{resummation}} -- the use of series techniques and calculus to obtain analytic results for the infinite series that appear when we consider an infinite number of splittings, and thus uncover the secrets of jet substructure.
%
We will also see \underline{renormalization} -- roughly, ``zooming in'' -- emerge as a useful lens for understanding the inner viscera of a jet as particles seem to split more and more as we zoom further and further in on jets.
%
Renormalization will give us additional tools to obtain finite results, and even allow us to recover the celebrated \gls{dglapeqn}.

% -----------------------------------
% Picturebook figure
% -----------------------------------
\reusefigure[t]{picturebook_jets}

@@

\epigraph{The typical high-energy event should consist in the production of a pair of virtual partons, each of which develops into a jet of physical hadrons.}{N. Cabibbo, G. Parisi, and M. Testa, \textit{Hadron Production in \(e^+ e^-\) Collisions} \cite{Cabibbo:1970mh}, 1970}

\epigraph{Continuing in this way, it should be possible to test the detailed predictions of QCD for production of any numbers of quarks and gluons, but always reinterpreting
these particles in terms of jets.}{G. Sterman and S. Weinberg, \textit{Jets from Quantum Chromodynamics} \cite{Sterman:1977wj}, 1977}

We have seen that partonic radiation is universal in the \glslink{soft-limit}{soft} and \glslink{collinear-limit}{collinear} limits, and that it can be described quantitatively as a \textit{splitting} process using \glspl{splittingfn} and \gls{angular-ordering}.
%
The particles emerging from these splittings can be grouped together into universal objects that we call jets.
%
We strongly believe that quarks and gluons are \textit{confined}, and cannot be directly observed.
%
Therefore, jets are a key feature of high-energy particle collisions that we must use to infer the presence of quarks and gluons.

\begin{definitionbox}{Tree of Partonic Splittings (The Partonic Cascade)}{tree}
    By applying philosophy of the previous section -- that differential cross sections in the collinear limit can be described in terms of \glslink{splittingfn}{\textit{splitting probabilities}} -- we model the formation of a jet as a series of splitting processes.
    %
    The resulting \glslink{tree}{\vocab{tree of partonic splittings}}, discussed in detail in \Sec{parton-shower}, is a construct of the soft-collinear limit of \gls{pqcd}.
    %
    The tree of partonic splittings, also called the partonic cascade, contains the full history and information of all of the splittings of a single parent parton (such as the kinematics and partonic flavor of any of its children due to splittings).
\end{definitionbox}

In order to provide useful information about our microscopic universe, jets must be defined in a way which is experimentally robust, theoretically well-defined, and produces results which can be compared to calculations in perturbation theory.
%
The study of the \textit{substructure} of jets can then be used to provide additional insights into the properties of the partons involved in high-energy collider processes.

@@

In this thesis, we use the \gls{akt} \cite{Cacciari:2008gp}, \gls{ca} \cite{Dokshitzer:1997in,Wobisch:1998wt}, and \gls{kt} \cite{Catani:1991hj,Catani:1993hr} jet algorithms, which are all \vocab{iterative \gls{reclustering} algorithms}, or \vocab{sequential recombination algorithms}:
%
algorithms which act on sets of outgoing momenta by recursively combining pairs of particles together until the algorithm reaches a termination condition.
%
When the algorithm terminates, it yields a set of jets associated with the input particle collision event.

The jet algorithms we use in this thesis (and nearly all jet algorithms) depend on a \vocab{jet radius} \Rjet, which determines the angular separation of particles within a jet.
%
Roughly, \(\Rjet\) is the radius of a circle in angular coordinates which determines a cone in which the jet constituents must lie;
%
if two particles are separated by an angle greater than \(2 \Rjet\), they will (usually) be associated with different jets.

\remark{}{
    The use of narrow jets, with \(\Rjet \ll 1\), justifies the use of the collinear limit, which will ease the calculation of observables which capture the substructure of jets.
    %
    In particular, the physics of narrow jets can be accurately characterized by the \glslink{collinear-limit}{collinear} splitting probabilities of \Sec{splitting-functions}, whose derivation assumed the collinear limit.
    %
    Furthermore, the use of \textit{central} jets, with rapidity \(y \sim 0\), will allow us to trade transverse momenta (which are more easily measured in hadronic collisions) for energy fractions (which appear in the theoretical description of jets through \glspl{splittingfn}).
}

@@

We have predicted the presence of collinear sprays of particles in scattering events the strong force and defined jets in order to capture their properties more precisely.
%
Now, we want to find a way to leverage jets to determine the fundamental properties of our universe.
%
In this thesis, we take the approach of studying the \textit{substructure} of jets.
%
The study of jet substructure has contributed to the deep connection between experimental and theoretical particle physics, with notable successes including precise determinations of the strong coupling constant \(\alpha_s\) and the on-going search for the precise value of the top quark mass \(m_t\) \cite{Procura:2022fid,Holguin:2022epo,Holguin:2023bjf,Pathak:2023tmy,Xiao:2024rol,Holguin:2024tkz}, whose value dictates the stability of the \glslink{sm}{Standard Model} vacuum and therefore the fate of the universe as we know it.

@@

\remark{}{
    The algorithmic formulation of \glspl{jet} was created in order to study jets as proxies of the quarks and gluons of QCD \cite{Sterman:1977wj}
    %
    in the algorithmic formulation, a jet is a coarse-grained degree of freedom roughly associated with a quark and its associated collinear radiation.
    %
    We note briefly, however, a complementary point of view that jets may also be thought of as a collinear enhancement of the \gls{energy-flow} (as in \Eq{energy-flow}) in a scattering event \cite{Basham:1978bw,Basham:1978zq,Basham:1979gh,Sveshnikov:1995vi,Cherzor:1997ak,Korchemsky:1997sy,Testa:1998gn,Korchemsky:1999kt}.
    %
    The latter idea is more connected to rigorous CFT methods \cite{Hofman:2008ar,Hatta:2012kn,Belitsky:2013bja,Belitsky:2013ofa,Belitsky:2014zha,Korchemsky:2015ssa,Goncalves:2014ffa,Farnsworth:2015hum,Hofman:2016awc,Kravchuk:2018htv,Kologlu:2019mfz,Chang:2020qpj,Korchemsky:2021htm,Caron-Huot:2022eqs,Chen:2023wah,Chicherin:2023gxt,Kologlu:2019mfz,Moult:2019vou,Chen:2020vvp,Chen:2024nyc} and may also be more intuitively connected to experimental results -- the energy flow \(\mathcal{E}(\hat{n})\) is roughly the energy passing through a detector at the angle \(\hat{n}\).
}

@@

% If you zoom far away from the constituents of a system (or zoom deeply in towards them), you're likely to end up at a renormalization fixed point.

@@

One of the first uses of renormalization -- apart from the study of \gls{qcd} itself -- was in the study of the presence and size of magnetic domains within magnetic materials, such as naturally-occurring iron.
%
Renormalization tells us how water molecules form snowflakes, and snowflakes form snow banks.
%
It is the theory of how structure and order forms from simplicity, and how disorder and chaos form from structure.

@@

\remark{}{
    The use of subjets within jets already hints at a fractal-like, self-similar structure of the partonic cascade.
    %
    An extremely concrete description of jets as fractals can be obtained using the technology of the \gls{emd};
    %
    see the discussion of \emph{correlation dimensions} in \Reff{Komiske:2019jim}.
}

If \gls{renormalization} is the physics, then \vocab{\gls{resummation}} is the math.
%
Theoretically, we have seen that jets can be imagined as the result of an arbitrary number of partonic splittings -- as few as zero and as many as infinity -- generated off of a parent quark, gluon, or other parton.
%
Resummation is generally the study of how to produce finite results from infinite series of this type -- especially when additional regularization is needed to produce a finite result.
%
Come to us in our hour of hope, it is exactly the mathematical framework we need.

\begin{definitionbox}{Resummation}{resummation}
    \glslink{resummation}{\vocab{Resummation}} is any method uused to turn (potentially divergent) series expansion into compact and often analytic mathematical expressions without any summation.
    %
    For our purposes, resummation will refer to the process of systematically reintroducing terms that have been previously expanded or neglected in perturbative expansions, in order to account for important contributions at higher orders or at specific scales.
\end{definitionbox}

Resummation is the technical segue we need to go from the conceptual developments of jets and renormalization to a quantitative study of jet substructure, and it is a necessary tool for removing the infinities and negative probabilities plaguing our the fixed order computations in \Sec{fixed-order-substructure}.
%
We perform simple resummed analyses in \Secs{ll-substructure-diagrams}{p2p-fragmentation} which leverage the recursive structure of the partonic cascade to obtain meaningful results at all orders in \(\alpha_s\) (unlike e.g. the analyses of \Sec{fixed-order-substructure}, which worked only to NLO).

Soon, we show that resumming of the infinity of partonic splittings reduces the seemingly massive complexity of the tree of partonic emissions into simple, compact formulae.
%
In fact, we characterize the fractal-like, psychadelic inner life of jets with simple exponential functions!
%
We say that the associated resummed quantities \vocab{exponentiate}.
%
A simple application of renormalization will then reveal what is called the \vocab{DGLAP Equation} -- a simple, physically intuitive result of scaling which reveals the deep heart of jet physics.

@@

\remark{}{
    The mathematical theory of resummation goes far beyond the paltry introduction of this chapter.
    %
    Borel resummation is particularly important for understanding the full formal structure of quantum field theory (appearing, for example, in the theory of renormalons and the non-perturbative structure).
    %
    Abel resummation, named after the mathematician whose quote appears at the beginning of this section, is another common resummation method for divergent series in mathematics.
    %
    For a more mathematical and complete presentation, I can recommend the classic mathematical textbook by Hardy \cite{Hardy_1991} and the classic physics articles of \Reffs{Dunne:2012ae}{Dorigoni:2014hea};
    %
    a full list of the many wonderful resources for resummation beyond these, is far beyond our scope.
}

% ==============================================
\section{The Cascade: A Universal Model of Jet Formation}
% ==============================================
\label{sec:parton-shower}

\epigraph{The picture then is one where matrix elements are likely to provide the better description of the main character of events, i.e. the topology of well separated jets, while parton showers should be better at describing the internal structure of these jets.}{J. Andr\'e and T. Sj\"ostrand, \textit{A matching of matrix elements and parton showers} \cite{Andre:1997vh}}

Finally, we are ready to describe the properties of the entire tree of partonic splittings.
%
As promised, we will compute the probability distribution for the properties of the partonic cascade by sewing together the results of consecutive partonic splittings, leading us to several deep and quite simple conclusions about the fractal nature of jets.
%
The simple description we provide in this chapter strictly holds only in \textit{strongly-ordered} regions of phase space discussed below, and is associated with results at \textit{double logarithmic accuracy} (see \Fig{dlla-plane}).
%
Nonetheless, the model we develop in this chapter facilitates a quantitative study of jet substructure.

We develop our technical description of the partonic cascade in \Sec{everything-on-shell}.
%
For those who wish to bypass the technical introduction and quickly access perturbative computations, here is the main conclusion of \Sec{everything-on-shell}:
\begin{answer}
    The technology of splitting functions developed in the previous chapter can be applied recursively in the soft-collinear limit.
    %
    This leads to a model of jets in which partons split recursively with a phase space probability distribution given by
    \begin{align}
        \splitxsec{\ell}{ij}
        =
        \frac{\alpha_s}{\pi}
        P_{ij \leftarrow \ell}(z_i)\,\frac{\dd z_i \, \dd \theta}{\theta}
        \,,
    \end{align}
    where \(z_i = E_i / E_\ell\) and \(\theta = \theta_{ij}\), exactly as in \Eq{dsigmatilde_definition} of \Sec{splitting-functions}.
\end{answer}
In \Secs{ll-substructure-diagrams}{p2p-fragmentation}, use our model of the partonic cascade to perform explicit calculations that will allow us to perform realistic calculations of jet substructure in \Chaps{grooming}{ewocs}.
%
In the next section, we will discuss re-clustering and low-energy pollution in preparation for a more realistic analysis of experimentally-obtained jets.

\begin{figure}[t!]
    \centering
    \includegraphics[width=0.7\textwidth]{example-image-a}
    \caption[
        Regions of the \(1/z\)-\(s\) plane associated with our approximations.
        %
        Adapted from \Reff{Collins:1984xc}
        %
        \sam{complete}
    ]
    {
        Regions of the \(1/x\)-\(s\) plane associated with the approximations use in this thesis.
        %
        Adapted from \underline{QCD and Collider Physics} \cite{Collins:1984xc}.
        %
        \sam{complete -- see page 132, fig 4.18}
    }
    \label{fig:dlla-plane}
\end{figure}

% ---------------------------------------
\subsection{Everything On-Shell: Strong Ordering of Successive Splittings}
% ---------------------------------------
\label{sec:everything-on-shell}

\epigraph{Taking matrix-element and phase-space factorization together, it follows that cross sections can be factorized. This allows for iteration of the approximation, as long as the measure of “softness” or “collinearity” remains appropriate. In this context, the requirement of an appropriate measure leads to the notion of strong ordering, which means that radiation of soft particles is yet softer and radiation of collinear particles is yet more collinear.}{\underline{A comprehensive guide to the physics and usage of \textsc{Pythia} 8.3} \cite{Bierlich:2022pfr}}

In this section, we discuss why and how we may sew together splitting functions in order to characterize the partonic cascade.
%
Recall from \Sec{splitting-functions} that splitting functions were defined in terms of amplitudes for the production of \textit{on-shell} final states, \(\mathcal{M}_{\ell^* \to i j}\).
%
In \Eq{}, we then wrote the full differential cross section for a process producing \(N\) particles, two of which were a collinear pair \((i_1, j_1)\), in a factorized form:
%
\begin{align}
    \dd \sigma_N
    =
    \dd \sigma_{N-1}
    \,
    \dsigmatilde_{i_1\,i_2}
    \,,
\end{align}
where \(\sigma_{N-1}\) is the ``differential cross section'' when the pair \((i_1,j_1)\) have been replaced by the off-shell \(\ell^*\), and is associated with the production of \(N-1\) particles, and \(\dsigmatilde_{i_1\, i_2}\) is the splitting distribution (as in \Prop{splitting-probability-main} of \Sec{splitting-functions}) associated with the splitting.

To continue recursively, we would like to write formulae such as, schematically,
\begin{align}
    \label{eq:cross-section-first-recursion}
    \dd\sigma_{N-1}
    =
    \dd \sigma_{N-2}
    \,
    \dsigmatilde_{i_2 \, j_2}
    \,,
\end{align}
where one of the collinear pair \((i_2, j_2)\) can potentially the \(\ell^*\) associated with the splitting \(\ell^* \to i_1 j_1\), and therefore
\begin{align}
    \dd \sigma_N
    =
    \dd \sigma_{N-2}
    \,
    \dsigmatilde_{i_1 \, j_1}
    \,
    \dsigmatilde_{i_2 \, j_2}
    \,.
\end{align}

However, in order apply all of the technology we introduced in \Chap{particles} to continue factorizing \(\dd\sigma_{N-1}\) as in \Eq{cross-section-first-recursion}, we need the assumptions of our propositions and lemmas to hold.
%
We will need to examine the features of angular ordering, the soft limit, and factorization in the presence of multiple splittings -- i.e. splittings in \(\dd \sigma_{N-1}\) given that we have already performed a splitting in factorizing \(\dd\sigma_N\).

Finally, we need \(\dd\sigma_{N-1}\) to describe the production of \(N-1\) \textit{on-shell} particles.
%
The fact that \(\ell^*\) is off-shell therefore provides an important technical complication.
